Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease

### 2023-05-10 doc/manuscript/translocation.qmd
Reduce viability methods and address Max's comments

Given observed yearly adult survival probabilities (from analysis of CMR data) and a baseline yearly recruitment probability $\omega$ of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
These six populations all had observed yearly adult survival greater than 0.5.
When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A).
In contrast, for the other six translocated populations, where yearly adult survival was substantially less than 0.5, our model predicted that it was unlikely that long-run growth rate in these populations was greater than 1, even when the probability of successful recruitment to the adult life stage was near 0.4 (@fig-viability A).
A sensitivity analysis of long-run growth rate $\lambda$ to yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$, recruitment probability $\omega$, and female fecundity *F* showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (@fig-viability-supp A).
Note that we assumed $\sigma_{A_R}$ was equivalent to the survival probability of translocated adults $\sigma_{A_T}$, except for three populations where we had \> 5 years of CMR data and recovering populations that together allowed us to estimate $\sigma_{A_R}$ and $\sigma_{A_T}$ separately (@fig-compare_surv_probs SI).
<!--# Consider condensing the above paragraph and providing details in SI. Next paragraph topic sentence could also provide a more direct statement of key results, e.g., something like "Even when incorporating demographic and environmental stochasticity, populations with high adult survival are likely to persist over a 50 year time horizon." -->

We then incorporated demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental stochasticity in the presence of Bd) into the model described above and used it to predict long-term population viability.
Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of 0 to 0.32; @fig-viability B).
This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment (@fig-viability B).
When we decreased $\omega$ from 0.3 to a conservative estimate of 0.15, 50 year extinction probability remained less than 0.5 for three populations and was 0.79 to 0.94 for three additional populations (@fig-viability B).
In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability-supp B; population 70550, surveyed for 16 years).

@@

The parameters in this model are yearly survival probability $\sigma_{\cdot}$ (the subscript "$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile $p_{J1}$., and the probability of a juvenile successfully recruiting to an adult $\omega$.

@@

###### Incorporating yearly variability in vital rates

###### Estimating model parameters

###### Model analysis and simulation

@@

![Results from the splined window analysis showing outlier windows for the difference in nucleotide diversity $\pi_{diff}$ (A) and *F~ST~* (B). Outlier regions for either $\pi_{diff}$ or *F~ST~* are shown in blue and outlier regions for both $\pi_{diff}$ and *F~ST~* are shown in red. (C) Magnified Contig19 from (A) showing two adjacent outlier windows for $\pi_{diff}$ 12.9Mb upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue line). (D) Magnified Contig8 from from (B) showing the *F~ST~* outlier window that includes the outlier SNPs TCF19 and VARS. This region of the genome contains 8 annotated genes known to occur in the extended MHC Class I and II regions.](figures/splinewindow_manhattan.png){#fig-spline-manhattan width="450"}

@@

@@

![Results from the exome capture selection study. (A) PCA calculated from binary SNPs showing genomic relationship of samples. Numeric labels and colors match those in @fig-allelefrequencies A. (B) qqplot showing observed and expected p-values for 148,307 SNPs and indels as calculated in GEMMA. Dashed line shows p-value that identifies outliers. (C) Manhattan plot showing p-values for each SNP as calculated by GEMMA. SNPs are sorted by genomic position and contigs are sorted by size. Red line shows p-value that identifies outliers. Outlier SNPs above this threshold are highlighted and labeled. Bold labels indicate at least one non-synonymous SNP is present in that gene.](figures/pca_qq_manhattan.png){#fig-selectionresults width="450"}

>>>>

Given observed yearly adult survival probabilities (from analysis of CMR data) and a baseline yearly recruitment probability $\omega$ of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
These six populations all had observed yearly adult survival greater than 0.5.
When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A).
<!--# Consider condensing the above paragraph and providing details in SI. Next paragraph topic sentence could also provide a more direct statement of key results, e.g., something like "Even when incorporating demographic and environmental stochasticity, populations with high adult survival are likely to persist over a 50 year time horizon."  Mark here:  Addressed both of these comments. -->

Even when incorporating demographic stochasticity and environmental stochasticity in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of 0 to 0.32; @fig-viability B).
This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment (@fig-viability B).
When we decreased $\omega$ from 0.3 to a conservative estimate of 0.15, 50-year extinction probability remained less than 0.5 for three populations and was 0.79 to 0.94 for three additional populations (@fig-viability B).
In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability-supp B; population 70550, surveyed for 16 years).

@@

The parameters in this model are yearly survival probability $\sigma_{\cdot}$ (the subscript "$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile $p_{J1}$, and the probability of a juvenile successfully recruiting to an adult $\omega$.

@@

After parameterizing our model with CMR-estimated adult frog survival probabilities and other known vital rates ([@tbl-param_values] : SI), we performed four analyses (described in detail in the SI).
First, we compute the long-run growth rate $\lambda$ for each of our 12 translocated populations to determine if the populations were deterministically predicted to growth or decline in the long-run.
Second, we compute the elasticity of $\lambda$ to model parameters, to quantify how much changes in adult survival, for example, affected the long-run growth rate.
Third, we included demographic and environmental stochasticity into our model and simulated the 50-year viability (i.e., 1 - extinction probability) of populations given an introduction of 40 adult individuals into an unoccupied habitat.
Finally, we fit our model to our longest translocation trajectory to confirm that our model could reasonably reproduce the observed recovery trajectories of MYL frogs following reintroductions.

@@

![Results from the splined window analysis showing outlier windows for the difference in nucleotide diversity $\pi_{diff}$ (A) and *F~ST~* (B). Outlier regions for either $\pi_{diff}$ or *F~ST~* are shown in blue and outlier regions for both $\pi_{diff}$ and *F~ST~* are shown in red. (C) Magnified Contig19 from (A) showing two adjacent outlier windows for $\pi_{diff}$ 12.9Mb upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue line). (D) Magnified Contig8 from from (B) showing the *F~ST~* outlier window that includes the outlier SNPs TCF19 and VARS. This region of the genome contains 8 annotated genes known to occur in the extended MHC Class I and II regions.](figures/splinewindow_manhattan.png){#fig-spline-manhattan}

@@

### Population viability modeling

##### Incorporating yearly variability in vital rates

##### Estimating model parameters

##### Model analysis and simulation

@@

![Results from the exome capture selection study. (A) PCA calculated from binary SNPs showing genomic relationship of samples. Numeric labels and colors match those in @fig-allelefrequencies A. (B) qqplot showing observed and expected p-values for 148,307 SNPs and indels as calculated in GEMMA. Dashed line shows p-value that identifies outliers. (C) Manhattan plot showing p-values for each SNP as calculated by GEMMA. SNPs are sorted by genomic position and contigs are sorted by size. Red line shows p-value that identifies outliers. Outlier SNPs above this threshold are highlighted and labeled. Bold labels indicate at least one non-synonymous SNP is present in that gene.](figures/pca_qq_manhattan.png){#fig-selectionresults}
